\huge \lm: \\ \centerline{Supporting Types for \ma}

### 2024-09-12 writing/IMEKOWC2024/presentation/presentation.tex

\title{\huge \lm: \\ \centerline{Supporting Types for \ma} }
\author[McBride, Nakov, Nordvall Forsberg et al]{\small Conor McBride$^{1}$, \underline{Georgi Nakov}$^{1}$, Fredrik Nordvall Forsberg$^{1}$,\\ Andr\'e Videla$^{1}$, Alistair Forbes$^{2}$, Keith Lines$^{2}$}
\institute[]{$^{1}$University of Strathclyde, UK\\$^{2}$National Physical Laboratory, UK}

\begin{document}

@@

\begin{frame}[fragile]{How does \lm Work?}
  \begin{itemize}[<+->]
  \item \lm is a \keyword{program transducer}: reads \ma code with formal comments, and outputs a modified version of the input.
    \medskip
  \item These formal comments are \keyword{directives} --- they start with \textcolor{commentgreen}{\texttt{\%<}}.
    \medskip
  \item Input the program:
   \begin{lstlisting}[xleftmargin=2em]
    %> rename n x !\tikzmark{ren}!
    n = 5;
    display(n);
   \end{lstlisting}
  \pause
  \mycallout<+>{ren}{this is an input directive to rename a variable}
\item Run \lm to get:
  \begin{lstlisting}[xleftmargin=2em]
%< LabMate 0.2.0.1 !\tikzmark{ver}!
%< renamed n x !\tikzmark{renr}!
x = 5;
display(x);
   \end{lstlisting}
  \pause
  \mycallout<+>{renr}{LabMate response to the input directive}
  \mycallout<+>{ver}{Mark the file as processed by LabMate}
\end{itemize}
\end{frame}

\begin{frame}[fragile]{Matrix Types}
  \begin{itemize}[<+->]
  \item \keyword{Matrices} feature heavily in \ma code.
    \medskip
  \item \lm supports \keyword{type annotations} for matrices:
\begin{lstlisting}[xleftmargin=0em, belowskip=-0.5em]
%> A :: [ 1 x 2 ] int  !\tikzmark{ann1}!
A = [ 3 4 ]
%> B :: [ 2 x 4 ] int  !\tikzmark{ann2}!
B = [ 1 1 1 1
      5 6 7 8 ]
C = A * B
\end{lstlisting}
  \pause
  \mycallout<+>{ann1}{type annotations at declaration of A}
  \mycallout<+>{ann2}{type annotations at declaration of B}
  \item We can ask for type information:
\begin{lstlisting}[xleftmargin=0em, belowskip=-0.5em]
%> typeof C !\tikzmark{query}!
%< C :: [Matrix 1 4 int]!\tikzmark{resp}!
\end{lstlisting}
 \pause
  \mycallout<+>{query}{query for the type of C}
  \mycalloutup<+>{resp}{LabMate can infer the dimensions}
  \item \ldots and can easily spot incompatible sizing:
\begin{lstlisting}[xleftmargin=0em]
D = B * A
%> typeof D
%< The expression D is!\tikzmark{bad}! quite a puzzle
\end{lstlisting}
    \pause
      \mycalloutup<+>{bad}{LabMate can point out an error with D}
\end{itemize}
\end{frame}

@@

A = 'hello'
%> typeof A
%< A :: string!\tikzmark{scope}!
\end{lstlisting}
   \pause
   \mycalloutup<+>{fromb}{LabMate infers type of A from the annotation on B}
   \mycalloutup<+>{scope}{tracks Matlab scope}
  \end{itemize}
\end{frame}

\begin{frame}[fragile]{Dimensions and Quantities}
  \begin{itemize}[<+->]
  \item \lm has support for arbitrary \keyword{quantities}.
    \begin{lstlisting}[xleftmargin=0em,belowskip=-1em]
%> dimensions V for Q over `Mass!\tikzmark{base}!, `Time
%> unit kg :: Q({!\tikzmark{unit1}! `Mass })!\tikzmark{unit}!
\end{lstlisting}
%\pause
   \mycalloutup<+>{base}{define some base set of dimensions}
   \mycalloutup<+>{unit}{and a canonical unit of measure}
   \mycalloutup<+>{unit1}{can be arbitrary group expression over V}
\pause
\begin{lstlisting}[xleftmargin=0em]
%<{
kg = 1; !\tikzmark{kg}!
%<}
\end{lstlisting}
\pause
   \mycallout<+>{kg}{this is a ``magic'' response that LabMate emits}
\pause
\item We can then use the \keyword{unit of measure} for quantities:
\begin{lstlisting}[xleftmargin=0em]
y = 5*kg !\tikzmark{q}!
%> typeof y
%< y :: Quantity (Enum [`Mass, `Time])
%<               {`Mass}
\end{lstlisting}
%  \pause
   \mycallout<+>{q}{turn a value of a dimensionless type into a quantity}
  \end{itemize}
\end{frame}

\begin{frame}[fragile]{Dimensional Consistency for Matrices}
  \begin{itemize}[<+->]
  \item Most \ma programs \keyword{do not work on uniform matrices}: type of the entry $e_{i, j}$ might depend on the indices $i$ and $j$.
    \medskip
  \item A common scenario when working with matrices of quantities:
%    $\begin{array}{ccccc}
% e_{1,1}  &        & e_{1,j} &        & e_{1,n} \\
%         &        &         &        &         \\
% e_{i,1}  &        & e_{i,j}\tikzmark{mat} &        & e_{i,n} \\
%         &        &         &        &         \\
% e_{m,1}  &        & e_{m,j} &        & e_{m,n}  \\
%\end{array}$
%    \mycallout<+>{mat}{the type might depend on i and j}
\begin{lstlisting}[xleftmargin=0em]
%> dimensions V for Q over `L, `M, `T
%> unit metre :: Q({ `L })
%> unit kg :: Q({ `M })
%> unit sec :: Q({ `T })
\end{lstlisting}
\medskip
\item Work in progress: \lm support for such matrices
\begin{lstlisting}[xleftmargin=0em]
% > A :: [ i <- [{} {`T}]
%        x j <- [{} {`L}]
%        ] Q({`M * j / i})
A = [ 2*kg       5*kg*metre
      3*kg/sec   4*kg*metre/sec ]
\end{lstlisting}
  \end{itemize}
\end{frame}

@@

>>>>

\title{\huge \lm: \\ \centerline{Supporting Types for \ma} }
\author[McBride, Nakov, Nordvall Forsberg et al]{\small Conor McBride$^{1}$, \underline{Georgi Nakov}$^{1}$, Fredrik Nordvall Forsberg$^{1}$,\\ Andr\'e Videla$^{1}$, Alistair Forbes$^{2}$, Keith Lines$^{2}$}
\institute[]{$^{1}$University of Strathclyde, UK\\$^{2}$National Physical Laboratory, UK}
\date{IMEKO 2024 XXIV World Congress\\August 27, 2024}
\begin{document}

@@

\begin{frame}[fragile]{How does \lm Work?}
  \begin{itemize}[<+->]
  \item \lm is a \keyword{program transducer}: reads \ma code with formal comments, and outputs a modified version of the input.
    \medskip
  \item These formal comments are \keyword{directives} --- they start with \textcolor{commentgreen}{\texttt{\%<}}.
    \medskip
  \item Input the program:
   \begin{lstlisting}[xleftmargin=2em]
    %> rename n x !\tikzmark{ren}!
    n = 5;
    display(n);
   \end{lstlisting}
  \mycallout<+>{ren}{this is an input directive to rename a variable}
  \item Run \lm to get:
   \begin{lstlisting}[xleftmargin=2em]
%< LabMate 0.2.0.1 !\tikzmark{ver}!
%< renamed n x !\tikzmark{renr}!
x = 5;
display(x);
   \end{lstlisting}
  %\pause
   \mycallout<+>{renr}{LabMate response to the input directive}
   \mycallout<+>{ver}{Mark the file as processed by LabMate}
\end{itemize}
\end{frame}

\begin{frame}[fragile]{Matrix Types}
  \begin{itemize}[<+->]
  \item \keyword{Matrices} feature heavily in \ma code.
    \medskip
  \item \lm supports \keyword{type annotations} for matrices:
    \begin{lstlisting}[xleftmargin=0em, belowskip=-0.5em]
%> A :: [ 1 x 2 ] int  !\tikzmark{ann1}!
A = [ 3 4 ]
%> B :: [ 2 x 4 ] int  !\tikzmark{ann2}!
B = [ 1 1 1 1
      5 6 7 8 ]
C = A * B
    \end{lstlisting}
    \mycallout<+>{ann1}{type annotations at declaration of A}
    \mycallout<+>{ann2}{type annotations at declaration of B}
  \item We can ask for type information:
    \begin{lstlisting}[xleftmargin=0em, belowskip=-0.5em]
%> typeof C !\tikzmark{query}!
%< C :: [Matrix 1 4 int]!\tikzmark{resp}!
    \end{lstlisting}
    \mycallout<+>{query}{query for the type of C}
    \mycalloutup<+>{resp}{LabMate can infer the dimensions}
  \item \ldots and can easily spot incompatible sizing:
    \begin{lstlisting}[xleftmargin=0em]
D = B * A
%> typeof D
%< The expression D is!\tikzmark{bad}! quite a puzzle
    \end{lstlisting}
    \mycalloutup<+>{bad}{LabMate can point out an error with D}
\end{itemize}
\end{frame}

@@

A = 'hello'
%> typeof A
%< A :: string!\tikzmark{scope}!
     \end{lstlisting}
     \mycalloutup<+>{fromb}{LabMate infers type of A from the annotation on B}
     \mycalloutup<+>{scope}{distinguishes and tracks Matlab scope}
  \end{itemize}
\end{frame}

\begin{frame}[fragile]{Dimensions and Quantities}
  \begin{itemize}[<+->]
  \item \lm has support for arbitrary \keyword{quantities}.
    \begin{lstlisting}[xleftmargin=0em,belowskip=-1em]
%> dimensions V for Q over `Mass!\tikzmark{base}!, `Time
%> unit kg :: Q({!\tikzmark{unit1}! `Mass })!\tikzmark{unit}!
    \end{lstlisting}
    \mycalloutup<+>{base}{define some base set of dimensions}
    \mycalloutup<+>{unit}{and a canonical unit of measure}
    \mycalloutup<+>{unit1}{can be arbitrary group expression over V}
    \pause
    \begin{lstlisting}[xleftmargin=0em]
%<{
kg = 1; !\tikzmark{kg}!
%<}
    \end{lstlisting}
    \pause
    \mycallout<+>{kg}{this is a ``magic'' response that LabMate emits}
    \pause
  \item We can then use the \keyword{unit of measure} for quantities:
    \begin{lstlisting}[xleftmargin=0em]
y = 5*kg !\tikzmark{q}!
%> typeof y
%< y :: Quantity (Enum [`Mass, `Time])
%<               {`Mass}
   \end{lstlisting}
   \mycallout<+>{q}{turn a value of a dimensionless type into a quantity}
  \end{itemize}
\end{frame}

\begin{frame}[fragile]{Dimensional Consistency for Matrices}
  \begin{itemize}[<+->]
  \item Most \ma programs \keyword{do not work on uniform matrices}: type of the entry $e_{i, j}$ might depend on the indices $i$ and $j$.
    \medskip
  \item A common scenario when working with matrices of quantities:
    \begin{lstlisting}[xleftmargin=0em]
%> dimensions V for Q over `L, `M, `T
%> unit metre :: Q({ `L })
%> unit kg :: Q({ `M })
%> unit sec :: Q({ `T })
    \end{lstlisting}
    \medskip
  \item Work in progress: \lm support for such matrices
    \begin{lstlisting}[xleftmargin=0em]
% > A :: [ i <-!\tikzmark{indi}! [{} {`T}]
%        x j <-!\tikzmark{indj}! [{} {`L}]
%        ] Q({`M * j / i})!\tikzmark{cell}!
A = [ 2*!\tikzmark{e00}!kg       5*!\tikzmark{e01}!kg*metre
      3*kg/sec   4*!\tikzmark{e11}!kg*metre/sec]
    \end{lstlisting}
    \mycalloutup<+>{cell}{The cell type Q depends on i and j}
    \mycalloutup<+>{indi}{Index i tells us something about the rows}
    \mycalloutup<+>{indj}{Index j disambiguates the columns}
    \mycalloutup<+>{e00}{:: Q(\{ `M * \{\} / \{\} \})}
    \mycalloutup<+>{e01}{:: Q(\{ `M * `L / \{\} \})}
    \mycalloutup<+>{e11}{:: Q(\{ `M * `L / `T \})}
  \end{itemize}
\end{frame}

@@

\begin{frame}
  \begin{center}
    \LARGE Thanks for the attention!\\
    \pause
    \vspace{1em}
    \Huge Questions?
  \end{center}
\end{frame}
